---
title: "Project 3"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(rlang)
library(magrittr)
library(colorspace)
library(jpeg)
library(cowplot)
library(broom)
library(kableExtra)
library(lubridate)
# don't write stuff to working directory
is.selfcontain <- TRUE

fetch.csv <- function(name, url, where = caller_env(), quiet = TRUE) {
    name <- ensym(name)
    if (!is.selfcontain) {
        fpath <- paste0("data/", name, ".csv")
        if (!file.exists(fpath)) {
            download.file(url, fpath)
        }
        env_poke(
            where, quo_name(name),
            if (quiet) {
                suppressMessages(readr::read_csv(fpath))
            } else {
                readr::read_csv(fpath)
            }
        )
    } else {
        if (quiet) {
            suppressMessages(readr::read_csv(url))
        } else {
            env_poke(where, quo_name(name), readr::read_csv(url))
        }
    }
}

moviedb.url.prefix <- "https://image.tmdb.org/t/p/w500/"
```

**Introduction:** 

We are looking at the [horror movies](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-11-01)
dataset from TidyTuesday.
This dataset contains metadata for horror movies from the [Movie Database](https://www.themoviedb.org/).
The whole dataset is ~36k records, but we will
exclude obscure movies that do not have a poster image available or not released yet.
We selected the top 10k remaining records by popularity for this analysis.

In this project we only used the year of the movie and the poster image URL from the dataset.

**Question:**

Are there particular pattern in the color scheme of horror movie posters?
How do the popularity of these patterns change over time?

**Approach:**

Firstly we need to download the poster image from the URL
specified in the dataset, and then extract the color features from the image.
This part is too computationally intensive to run in this document and
was implemented in Rust as an R extension, compiled and ran on the Edu Pod.

We wrote extractors that extracts the following into a CSV file:

- Brightness, by averaging, in the range [0, 255]. In RGB space.
- Contrast, by RMSD. Output is in the range [0, 255]. In RGB space.
- Dominant colors, by an adaptive k-means clustering algorithm
  which tries to use between 3 and 8 color clusters.
  The optimal number of clusters is selected using the silhouette score (within sum of squares),
  when adding another cluster reduces the score by less than 30% the algorithm stops.
  Each trial is repeated 8 times with different initializations and the best result is selected.
  The clusters are then sorted by size and the top 3 clusters are output. This is done in
  LAB space because the LAB space is more "sphere-like", with two components mapping to
  hue making it easier to interpret as color "tendencies".

If you are not familiar with the LAB colorspace, it has three components:

- L: Lightness, in the range [0, 100].
- A: Green to red, in the range [-90, +90].
- B: Blue to yellow, in the range [-90, +90].

The download and extractions took about 20/5 minutes each on Edu Pod using 32 threads,
I included an output CSV file on GitHub gist and we will use that for the analysis.

If you want to generate the CSV file yourself,
a sample pipeline is provided at the end of this document.

After we loaded everything into R,
we run a second k-means on the extracted features.
Then we will interpret the clusters based on their deviation from the mean
in each feature.

Lastly, we plot the % of movies in each cluster over time and
observe any trends. This is a multiple time-series so we will
use a faceted line plot.

```{r load_data}
# Load your dataset here
fetch.csv(horror_movies,
    "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-11-01/horror_movies.csv",
    where = global_env()
)
horror_movies %<>%
    # filter out movies with no poster or too low in popularity
    filter(status == "Released") %>%
    filter(!is.na(poster_path)) %>%
    arrange(desc(popularity)) %>%
    slice_head(n = 10000) %>%
    mutate(
        filename = str_remove(poster_path, "^/"),
    )

# load directly from processed output
# image_features <- read_csv("data/horror_movies_image_features.csv") %>%
#    distinct(filename, .keep_all = TRUE)

# data file with extracted features
fetch.csv(image_features,
    "https://gist.githubusercontent.com/eternal-flame-AD/74a23ff3086a3815ebd8216f9ba37cbe/raw/4911db6ec7ce4feec9a8ebe5c4dec9affaa11cac/horror_movies_image_features.csv",
    where = global_env()
)
image_features %<>%
    distinct(filename, .keep_all = TRUE)

# join the two datasets
horror_movies.joined <- horror_movies %>%
    inner_join(image_features, by = "filename", keep = FALSE)
```

**Analysis:**

To limit the scope of interpretation, 
we will only include the top 2 dominant colors
along with the brightness and contrast features
for clustering.

Cluster number of 6 is selected based on the elbow method
on the within sum of squares metric.

```{r do_kmeans}
do.kmeans <- function(centers) {
    set.seed(123)

    horror_movies.joined %>%
        select(
            # first dominant color
            starts_with("color_cluster_0"),
            # second dominant color
            starts_with("color_cluster_1"),
            starts_with("brightness_"),
            contrast_avg
        ) %>%
        scale() %>%
        kmeans(
            # some variables are too similar
            # for the default algorithm
            algorithm = "Lloyd",
            centers = centers,
            iter.max = 500,
            nstart = 25
        )
}

km_fit <- do.kmeans(6)
```

First we will plot the deviation
in each variable
from overall center for each cluster.

This will give us an idea of "where"
each cluster is and how they each differ from others.

```{r kmean_fit}
km_fit$centers %>%
    as_tibble() %>%
    rownames_to_column("cluster") %>%
    pivot_longer(-cluster, names_to = "parameter", values_to = "value") %>%
    mutate(
        parameter = factor(parameter, levels = c(
            "brightness_r",
            "brightness_g",
            "brightness_b",
            "brightness_avg",
            "contrast_avg",
            "color_cluster_0_l",
            "color_cluster_0_a",
            "color_cluster_0_b",
            "color_cluster_1_l",
            "color_cluster_1_a",
            "color_cluster_1_b"
        ))
    ) %>%
    ggplot(aes(fct_rev(parameter), value)) +
    geom_col(aes(fill = cluster), position = "dodge") +
    coord_flip() +
    facet_wrap(~cluster) +
    labs(
        y = "SDs from center",
        x = NULL,
        title = "Deviation from mean for each cluster by variable"
    ) +
    theme_bw() +
    theme(
        legend.position = "none"
    )
```

We interpreted this plot as follows:

```{r centroid_labeling, echo = FALSE}
centroid.names <- c(
    "1" = "-Br-Ct",
    "2" = "+Ct 1+L-a",
    "3" = "Nil",
    "4" = "+Br 0+a+b",
    "5" = "1+a+b",
    "6" = "+Br+Ct 0+L"
)
centroid.as.factors <- function(idx) {
    idx %>%
        as.character() %>%
        factor(levels = names(centroid.names), labels = centroid.names) %>%
        fct_relevel("Nil")
}

centroid.names.human <- c(
    "1" = "Just Dark",
    "2" = "Bright on Dark",
    "3" = "No significance",
    "4" = "Bright Red/Yellow",
    "5" = "Dark with Warm Secondary Color",
    "6" = "Bright and Vibrant"
)
centroid.def <- data.frame(
    idx = names(centroid.names),
    name = centroid.names,
    human = centroid.names.human
)
humanize.cluster <- function(fct) {
    map_chr(fct, ~ {
        centroid.def %>%
            filter(idx == .x | name == .x) %>%
            pull(human)
    })
}
```

```{r clustered_data}
# join the cluster labels to the original dataset
horror_movies.clustered <- km_fit %>%
    augment(horror_movies.joined) %>%
    mutate(
        cluster = centroid.as.factors(.cluster)
    )
```

```{r cluster_table, echo = FALSE}
horror_movies.clustered %>%
    count(cluster) %>%
    mutate(
        prop = n / sum(n)
    ) %>%
    mutate(
        name = as.character(cluster)
    ) %>%
    inner_join(
        centroid.def
    ) %>%
    arrange(idx) %>%
    select(idx, name, human, n, prop) %>%
    knitr::kable(
        caption = "Cluster Naming and Interpreting Cluster Characteristics",
        col.names = c("Cluster", "Shorthand Name", "Description", "Count", "Proportion")
    ) %>%
    kable_styling(
        full_width = FALSE,
        latex_options = c("striped", "hold_position")
    )
```

To illustrate the clusters,
and visually confirm our interpretation,
we will select 12 posters from each cluster
and plot them on a tile plot.

```{r showcase}
if (require(ggpattern, quietly = TRUE)) {
    showcase <- horror_movies.clustered %>%
        group_by(cluster) %>%
        mutate(
            ord = row_number()
        ) %>%
        filter(ord <= 12)

    showcase %>%
        ggplot(
            aes(x = cluster, y = ord)
        ) +
        geom_tile_pattern(
            aes(pattern_filename = filename),
            pattern = "image",
            pattern_type = "expand"
        ) +
        coord_flip() +
        scale_pattern_filename_manual(
            values = set_names(showcase$url, showcase$filename)
        ) +
        labs(
            x = NULL, y = NULL,
            title = "Showcase for Clustered Movie Posters"
        ) +
        scale_x_discrete(
            expand = c(0, 0),
            labels = humanize.cluster
        ) +
        scale_y_continuous(
            expand = c(0, 0)
        ) +
        theme(
            axis.title.x = element_blank(),
            axis.text.x = element_blank(),
            axis.ticks.x = element_blank(),
            axis.text.y = element_text(size = 8),
            axis.ticks.y = element_blank(),
            legend.position = "none"
        )

    ggsave("showcase.png")
} else {
    warning("ggpattern not installed, skipping showcase")
}
```

Visually inspecting the output it does seem that the
interpretation of centroids is reasonable.

Thus we move on to plot the popularity of each type
over year.

```{r year_stacked_col, fig.width = 7, fig.height = 7}
horror_movies.clustered %>%
    mutate(year = year(release_date)) %>%
    select(cluster, year) %>%
    arrange(year) %>%
    # filter year range and count up by cluster and year
    filter(year >= 1970) %>%
    count(cluster, year) %>%
    # compute proportions
    group_by(year) %>%
    mutate(prop = n / sum(n)) %>%
    ggplot(aes(year, prop, color = cluster)) +
    geom_line() +
    scale_y_continuous(labels = scales::percent) +
    facet_wrap(~cluster) +
    scale_color_discrete(
        labels = humanize.cluster,
        guide = guide_legend(
            title = "Cluster Interpretation",
            title.position = "top",
            label.hjust = 0,
            nrow = 2,
            byrow = TRUE
        )
    ) +
    labs(
        x = "Year",
        y = NULL,
        title = "Popularity of Horror Movie Poster Types Over Time"
    ) +
    theme_bw() +
    theme(
        plot.background = element_rect(
            fill = "white",
            color = "grey70", linewidth = 2
        ),
        legend.position = "bottom",
        legend.margin = margin(t = 0, r = 0, b = 0, l = -1.5, unit = "cm"),
    )
```

```{r save_plot3, echo = FALSE}
ggsave("plot3.png", width = 7, height = 7)
```

**Discussion:**

Within the time range of 1970 to 2022,
We made the following observations:

- The proportion of "Just Dark" types of posters
  has been increasing over the years.
  It was very rare in the 70s but has grown to
  about 40% after 2020.
- The "Bright on Dark" type
  used to be popular but has been declining
  in population ever since.
- The "Bright Red/Yellow" pattern is staying
  rare at <10% from the 70s and staying at ~5%
  in the 21st century.
- The "Dark with Warm Secondary Color" has a
  steady popularity of around 20% for the
  past 50 years.
- The "Bright and Vibrant" cluster has been
  popular in the 70s at a rate of about 30% while
  afterwards it reduced to about 10%.

**Supplementary Information:**

Image download and extraction pipeline:

A source tarball of the image processor is available [here](https://mfstorstd02.blob.core.windows.net/sec/20232_SDS375/imaging-lib.tar.gz?sv=2021-10-04&st=2023-04-11T02%3A34%3A41Z&se=2023-05-12T02%3A34%3A00Z&sr=b&sp=r&sig=z0VPcfEpBTrXcqW0qOBgHuqY25OoSQclRDyuLqdi3IQ%3D).
Extract the tarball in the same directory as this document and run `load.extractor.api()` as defined below.
Then run `imaging.download()` to download the images and `imaging.extract()` to extract the features.

```{r pipeline_demo}
# workflow for downloading and extracting features
load.extractor.api <- function(loglevel = "INFO") {
    if (exists(".imaging.api.loaded")) {
        return()
    }
    .imaging.api.loaded <<- TRUE

    # generate input for image downloader/feature extractor
    if (!file.exists("data/horror_movies_urls_in.csv")) {
        horror_movies.urls <- horror_movies %>%
            transmute(
                url = paste0("https://image.tmdb.org/t/p/w500", poster_path),
                filename = filename
            )
        write_csv(horror_movies.urls, "data/horror_movies_urls_in.csv")
    }
    if (!file.exists("imaging-lib/target/release/libimaging_lib.so")) {
        message("Compiling Rust library...")
        system2(
            "bash",
            c("-c", shQuote("cd imaging-lib && cargo build --release"))
        )
    }
    dyn.load("imaging-lib/target/release/libimaging_lib.so")
    .Call("imaging_lib_init", loglevel)
    imaging.download <<- function(input_filename = "data/horror_movies_urls_in.csv",
                                  output_filename = "data/horror_movies_images_out.csv",
                                  output_zip_filename = "data/images.zip",
                                  limit = 0L,
                                  overwrite = FALSE) {
        ret <- .Call(
            "imaging_lib_download",
            input_filename, output_filename,
            output_zip_filename, limit, overwrite
        )
        if (is.character(ret)) {
            message(ret)
        }
    }
    imaging.extract <<- function(input_filename = "data/horror_movies_images_out.csv",
                                 output_filename = "data/horror_movies_image_features.csv",
                                 input_zip_filename = "data/images.zip",
                                 num_threads = 32L,
                                 limit = 0L,
                                 overwrite = FALSE,
                                 verbose = FALSE) {
        ret <- .Call(
            "imaging_lib_extract",
            input_filename, output_filename,
            input_zip_filename,
            num_threads, limit,
            overwrite, verbose
        )
        if (is.character(ret) && ret != "") {
            message(ret)
        }
    }
}

demo.pipeline <- function(limit = 3, n_threads = 1) {
    tmpf1 <- tempfile()
    tmpf2 <- tempfile()
    tmpf3 <- tempfile()
    imaging.download(
        input_filename = "data/horror_movies_urls_in.csv",
        output_filename = tmpf1,
        output_zip_filename = tmpf2,
        limit = limit
    )
    imaging.extract(
        input_filename = tmpf1,
        output_filename = tmpf3,
        input_zip_filename = tmpf2,
        num_threads = n_threads,
        limit = limit,
        verbose = TRUE
    )
    unlink(c(tmpf1, tmpf2, tmpf3))
}
if (!is.selfcontain) {
    tryCatch(
        {
            load.extractor.api()
            demo.pipeline()
        },
        error = function(e) {
            message("Failed to load Rust library, skipping demo pipeline.")
        }
    )
}
```